---
author: "Ira Cooke"
output: ioslides_presentation
---

# Exploring data with ggplot2

## Classic plots vs ggplot { .build }
  
### `plot`, `barplot`, `hist`
  
- One function per plot style
- Limited ability to modify the plot (eg colours) depending on plot type
- Inconsistent data requirements (eg matrix vs data frame, tall vs wide)


### `ggplot2`

- One package many, many, many plot styles
- Many aspects of plot appearance can be made to depend on the data
- Data should generally be tall rather than wide. Use `melt` and `cast` to transform.


## Grammar of Graphics { .build }

ggplot is an implementation of the [grammar of graphics](http://vita.had.co.nz/papers/layered-grammar.pdf)

>- Data
>- Geometric objects (points, lines, areas)
>- Aesthetic mappings between data and objects
>- Statistical transformations (binning, aggregating)
>- Position adjustment (stack, dodge, jitter)
>- Coordinate system and scale
>- Annotations (title, legend etc)
>- Faceting (representing data across multiple plots)

## Layers in ggplot {.build}

Each layer in ggplot has at least `data`, `geom` (geometric object), `stat` (statistical transform) and `aes` (aesthetic)

The `ggplot` function sets up a plot's baseline defaults

```{r,eval=FALSE}
p<-ggplot(data,aes(x=someval,y=someval))
```

Add layers with the `+` operator

```{r,eval=FALSE}
p+geom_point()
```

This layers inherits `data` and `aes` from the base plot and has a default `stat`

## A simple scatterplot {.build}


```{r,eval=FALSE}
library(ggplot2)
data(diamonds)
### <b>
p<-ggplot(diamonds)
p+geom_point(aes(x=carat,y=price))
### </b>
```

```{r, echo=FALSE}
library(ggplot2)
data(diamonds)
### <b>
ggplot(diamonds)+geom_point(aes(x=carat,y=price))
### </b>
```

## Add another aesthetic mapping {.build}

```{r, eval=FALSE}
p<-ggplot(diamonds)
### <b>
p+geom_point(aes(x=carat,y=price,color=clarity))
### </b>
```

```{r, echo=FALSE}
library(ggplot2)
data(diamonds)
ggplot(diamonds,aes(x=carat,y=price))+geom_point(aes(color=clarity))
```

## Available aesthetics {.build}

Aesthetics for `geom_point` are;

- x
- y
- alpha
- colour
- fill
- shape
- size

Find available aesthetics at [http://docs.ggplot2.org/current/](http://docs.ggplot2.org/current/). 

<div class="notes">
Aesthetics map properties of geometric object object (`geom`) to the data. Each `geom` has a different set of available properties. 
</div>


## Adding another layer {.build}

<div class="notes">
Each layer in ggplot has its own `data`,`aes`,`geom` and `stat`. These are all required, but ggplot has a hierarchical defaults system which means that in practice they are often inferred. 
</div>

```{r, eval=FALSE}
p<-ggplot(diamonds,aes(x=carat,y=price))
p+geom_point(aes(color=clarity))
### <b>
p+geom_smooth()
### </b>
```

```{r, echo=FALSE}
library(ggplot2)
data(diamonds)
ggplot(data=diamonds,aes(x=carat,y=price))+geom_point(aes(color=clarity))+stat_smooth(method="gam",formula=y ~ s(x, bs = "cs"))
```

## geom and stat objects are often complementary {.build}

For example `geom_smooth()` has `stat_smooth()` as its default `stat`

A `stat` adds new columns to the data that are used by the geom. For example `stat_smooth()` creates;

- `y` predicted value
- `ymin` lower pointwise confidence interval around the mean
- `ymax` upper pointwise confidence interval around the mean
- `se` standard error

These can all be accessed for aesthetic mapping if needed.

## Another `stat` example {.build}

```{r, eval=FALSE}
p<-ggplot(data=diamonds,aes(x=carat,y=price))
### <b>
p<-p+stat_density2d(aes(color=..level..))
### </b>
p+scale_color_continuous("Density")
```

```{r, echo=FALSE}
library(ggplot2)
p<-ggplot(data=diamonds,aes(x=carat,y=price))
p+stat_density2d(aes(color=log(..level..)))+scale_color_continuous("Density")
```

## Where does `..level..` come from?

```{r,echo=TRUE}
help(stat_contour)
```

...

<h3 style="color: rgb(40%, 40%, 40%);font-family: monospace;font-size: large">Value</h3>

<p style="color: black;font-family: monospace;font-size: medium">A data frame with additional column:
</p>
<table style="color: black;font-family: monospace;font-size: medium" summary="R valueblock">
<tr style="color: black;font-family: monospace;font-size: medium" valign="top"><td><code>level </code></td>
<td>
&nbsp;height of contour
</td></tr>
</table>

...

## Facetting {.build}

```{r, eval=FALSE}
p<-ggplot(data=diamonds,aes(x=carat,y=price))
p<-p+geom_point(aes(color=cut))
### <b>
p+facet_grid(color~cut)
### </b>
```

```{r, echo=FALSE}
library(ggplot2)
p<-ggplot(data=diamonds,aes(x=carat,y=price))
p<-p+geom_point(aes(color=cut))
p+facet_grid(color~cut)
```

## Reshaping data {.smaller}

Use the `reshape2` package to easily transform data for `ggplot`

```{r}
d<-read.csv("mqdata.csv")
head(d)
```

## Melt data from wide to tall {.smaller .build}

```{r}
library(reshape2)
md=melt(d,id.vars=c("Protein","nmiss"),variable.name="Sample",value.name="Intensity")
head(md)
```

## Cast from tall to wide {.smaller .build}

```{r}
cmd=dcast(md,Protein+nmiss~Sample,value.var='Intensity')
head(cmd)
```

## Visualise Missingness

```{r,eval=FALSE}
p<-ggplot(md,aes(x=as.factor(nmiss),y=log2(Intensity)))
p+geom_boxplot()+facet_wrap(~Sample)
```

```{r,echo=FALSE}
options(warn=-1)
suppressWarnings(ggplot(md,aes(x=as.factor(nmiss),y=log2(Intensity)))+geom_boxplot()+facet_wrap(~Sample))
```



